\section{Related work}



\shortpara{Representation alignment.}
Similarities between representations of
different neural networks are investigated in \citep{laakso2000content, li2016convergentlearningdifferentneural, wang2018understandinglearningrepresentationsextent, bansal2021revisitingmodelstitchingcompare, huh2024platonicrepresentationhypothesis, wolfram2025layers, li-etal-2024-vision-language}. Methods based on CCA \citep{morcos2018insightsrepresentationalsimilarityneural}, SVCCA, \citep{raghu2017svccasingularvectorcanonical}, CKA \citep{kornblith2019similarity, maniparambil2024vision}, ICA \citep{yamagiwa2023discoveringuniversalgeometryembeddings}, time-series \citep{mistry2023comparative}, and GUIs \citep{heimerl2019embComp} have been used to compare embeddings from different subspaces.  
\citep{maiorca2024latentspacetranslationsemantic, moschella2023relativerepresentationsenablezeroshot, moayeri2023texttoconceptandbackcrossmodel, tian2019latent, norelli2023asifcoupleddataturns} harness representation similarity for zero-shot stitching, substitution, domain transfer, and multimodal adaptation.  All rely on some amount of paired data, which is difficult  
to reduce \citep{cannistraci2023bootstrappingparallelanchorsrelative}.  Our method does not just measure similarity, we learn how to \emph{translate} representations across spaces without any paired data.


\shortpara{Optimal transport.}  The problem of unsupervised optimal transport has been studied for image style transfer \citep{huang2018multimodalunsupervisedimagetoimagetranslation, liu2018unsupervisedimagetoimagetranslationnetworks,zhu2020unpairedimagetoimagetranslationusing}, word translation \citep{xing2015normalized, conneau2018wordtranslationparalleldata, grave2018unsupervisedalignmentembeddingswasserstein, chen2018unsupervisedmultilingualwordembeddings, joulin-etal-2018-loss}, and natural language sequence translation \citep{raviknight2011deciphering, lample2018unsupervisedmachinetranslationusing, alvarezmelis2018gromovwassersteinalignmentwordembedding, artetxe2018unsupervisedneuralmachinetranslation, yang2018unsupervisedneuralmachinetranslation, artetxe-etal-2018-robust}.
Our method builds on these works, which often employ a combination of cycle-consistency and adversarial loss. Importantly, unlike prior word and sequence translation methods, multiple representations of the same input (e.g., heavily overlapping word vocabularies) are unavailable in our setting. \citep{schnaus2025itsblindmatchvisionlanguage} proposes a solver for matching small sets of embeddings between different vision-language models.
Our method goes well beyond matching by taking unknown embeddings and \emph{generating} matching embeddings in the space of another model.

\shortpara{Embedding inversion.} An emerging line of research investigates decoding text from language model embeddings \citep{song2020informationleakageembeddingmodels, li2023sentenceembeddingleaksinformation, morris2023textembeddingsrevealalmost} and outputs \citep{morris2023languagemodelinversion, carlini2024stealingproductionlanguagemodel, zhang2024extractingpromptsinvertingllm}.  
\atk{} helps apply these to unknown embeddings, without an encoder or paired data, by translating them to the space of a known model. 

\shortpara{Bridging modality gaps.} Previous work has noted an inherent ``gap” between image- and text-based models \citep{liang2022mindgapunderstandingmodality} and proposed various ways to unify the modalities \citep{song2023bridgegapmodalitiescomprehensive}. Some approaches feed image embeddings directly into language models \citep{koh2023fromage, wang2023visionllm, dong2024dreamllmsynergisticmultimodalcomprehension, liu2023visualinstructiontuning}, while others generate captions from image embeddings \citep{mokady2021clipcap} or even from text embeddings themselves \citep{morris2023textembeddingsrevealalmost}. \citep{girdhar2023imagebind} introduces a shared embedding space that integrates inputs from multiple modalities, including text, audio, and vision. In contrast, our post-hoc approach directly translates between representations and complements these systems by enabling inputs from a wide variety of embedding models.